---
title: "Week 7: Practice function making with rasters"
format: html
editor: 
  markdown: 
    wrap: 72
---

Discussion here: https://eds-223-geospatial.github.io/course-materials/discussions/week7-discussion.html

```{r}
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
```

# 1. Get Started

```{r}
magpie <- read_csv(here::here("week-7", "data", "magpie_obvs.csv"))
tule_elk <- read_csv(here::here("week-7","data", "tule_elk_obvs.csv"))

bioclim_dir <- here::here("week-7", "data", "climate", "wc2.1_2.5m")
bioclim <- list.files(bioclim_dir, pattern = glob2rx("*.tif$"), full.names = TRUE)
bioclim_sort <- bioclim[
  # Sort filepaths based on numeric suffix
  order(
    # Extract numeric suffix of filenames and convert to numeric
    as.numeric(gsub(".*_(\\d+)\\.tif$", "\\1", bioclim))
    )
  ]
bioclim_rast <- rast(bioclim_sort)
```

# 2. Name raster layers

Update the bioclim_rast layer names to match the Bioclimatic variables using names()

```{r}
names(bioclim_rast) # yikes
names(bioclim) # null

variables <- c("annualMeanTemp", "meanDiurnalRange", "isothermality", "tempSeasonality", "maxTempWarmMonth", "maxTempColdMonth", "tempAnnualRange", "meanTempWetQ", "meanTempDryQ", "meanTempWarmQ", "meanTempColdQ", "annualPrecip", "precipWetMonth", "precipDryMonth", "precipSeasonality", "precipWetQ", "precipDryQ", "precipWarmQ", "precipColdQ")
names(bioclim_rast) <- variables
```

# 3. Find geographic extent of species occurrence and crop raster

Use st_as_sf to convert and find the geographic extent of species occurrence data magpie using st_bbox()
Crop bioclim_rast to match the extent of species occurrence data magpie

```{r}
names(magpie)
magpie_sf <- magpie %>% 
  rename(long = longitude, # prettier names
         lat = latitude) %>% # prettier names
  drop_na(long, lat) %>% # drop long ones
  st_as_sf(coords = c("long", "lat"), crs = 4326) # make sf

# Obtain geographic extent/bounding box of species occurrences
magpie_bbox <- st_bbox(magpie_sf)

# Crop raster to match geographic extent of species occurrences
bioclim_crop <- crop(bioclim_rast, magpie_bbox)
```

4. Extract points from species occurrence

Extract points from the cropped bioclim_rast for each occurrence in magpie

```{r}
# Extract points from raster for all species occurrences
bioClim_pts <- as_tibble(terra::extract(bioclim_crop, magpie_sf)) # function terra::extract!
```

5. Create background values and extract points

Create “background values” by generating random sample points from the cropped bioclim_rast
Extract points from cropped bioclim_rast for the background values

This is common in species distribution modeling (e.g., MaxEnt or GLMs) where you need pseudo-absence or background points

```{r}
# Generate random sample points from raster
random_pts <- spatSample( # terra function that randomly samples cells from a raster
  bioclim_crop[["annualMeanTemp"]], # selects one raster layer (the annual mean temperature layer) from a multilayer raster object called bioclim_crop
                         na.rm = TRUE, # Random sample non-NA cells
                         size = (nrow(magpie) * 2), # creating a sample size based on the magpie, I want twice as many rows as the magpie
                         as.points = TRUE) %>% # Return SpatVector
  st_as_sf() # make it spatial

# Extract points from raster for background values
bioClim_random_pts <- as_tibble(terra::extract(bioclim_crop, random_pts)) # taking the random spatial points we just generated and extracting the environmental values at those locations from the raster stack
```

6. Plot species climate niche

```{r}
map_1 <- tm_shape(bioclim_crop[["annualPrecip"]]) +
  tm_raster(col.scale = tm_scale_continuous(values = "Blues"), 
            col.legend = tm_legend(title = "Annual precipitation")) +
  tm_shape(magpie_sf) +
  tm_dots(fill = "#3a5a40", size = 0.15) +
  tm_layout(legend.position = tm_pos_in("center", "bottom"),
            legend.orientation = "landscape",
            legend.bg.color = "white")

map_2 <- tm_shape(bioclim_crop[["annualMeanTemp"]]) +
  tm_raster(col.scale = tm_scale_continuous(values = "-RdYlBu"), 
            col.legend = tm_legend(title = "Annual mean temperature")) +
  tm_shape(magpie_sf) +
  tm_dots(fill = "#3a5a40", size = 0.15) +
  tm_layout(legend.position = tm_pos_in("center", "bottom"),
            legend.orientation = "landscape",
            legend.bg.color = "white")

tmap_arrange(map_1, map_2)
```

```{r}
plot_1 <- ggplot(data = bioClim_pts, aes(x = annualPrecip, y = annualMeanTemp)) +
  geom_point(shape = 16, color = "#3a5a40") +
  labs(x = "Annual precipitation",
       y = "Annual mean temperature", 
       title = "Species climate niche") +
  theme_bw()

plot_2 <- ggplot(data = bioClim_random_pts, aes(x = annualPrecip, y = annualMeanTemp)) +
  geom_point(shape = 16) +
  labs(x = "Annual precipitation",
       y = element_blank(), 
       title = "Background climate") +
  theme_bw()

plot_1 + plot_2
```


7. Create generalizable workflow and apply

Modify previous steps into a generalizable workflow for other species occurrence data
Try your new function with species occurrence data tule_elk

```{r}
climate_envelope <- function(clim_rast, # need to input a climate raster
                             clim_var1, # specifically the name of the layer 
                             clim_var2, # climate variable 2
                             occurences, # df with occurrences of the species and a column "long" and "lat" wiht longitude and latitude
                             species_name){ # just the name of the species, literally just a string, for storing the plot and for the ggplot later on
  
  # clean the name of the species snake_case
  species_name_lower <- species_name %>% 
    str_to_lower() %>% 
    str_replace_all(" ", "_")
  
  
  occurences_sf <- occurences %>% 
    rename(long = longitude, # ensure in your input there is a column long
           lat = latitude) %>% # ensure in your input there is a column lat
    drop_na(long) %>% # no na's in the coordinates
    st_as_sf(coords = c("long", "lat"), crs = 4326) # make it spatial

#....................crop to the occurrences and pull the points.....................
  # now we make the bbox to prepare to crop
  occurences_bbox <- st_bbox(occurences_sf) # make it a bbox
  
  # now crop to the bbox
  clim_crop <- crop(clim_rast, occurences_bbox) # crop the climate raster to the bbox

  # pull the points from the cropped dataframe 
  clim_pts <- as_tibble(extract(clim_crop, occurences_sf)) 
  
#.............randomly sample for background climate.............
  random_pts <- spatSample(clim_rast[[clim_var1]],
                           na.rm = TRUE,
                           size = (nrow(occurences) * 2),
                           as.points = TRUE) %>%
    st_as_sf()

  clim_random_pts <- as_tibble(terra::extract(clim_crop, random_pts))

#.........................build the maps......................... 
  map_1 <- tm_shape(clim_crop[[clim_var1]]) +
    tm_raster(col.scale = tm_scale_continuous(values = "Blues")) +
    tm_shape(occurences_sf) +
    tm_dots(fill = "#3a5a40", size = 0.15) +
    tm_layout(legend.position = tm_pos_in("center", "bottom"),
              legend.orientation = "landscape",
              legend.bg.color = "white")

  map_2 <- tm_shape(clim_crop[[clim_var2]]) +
    tm_raster(col.scale = tm_scale_continuous(values = "-RdYlBu")) +
    tm_shape(occurences_sf) +
    tm_dots(fill = "#3a5a40", size = 0.15) +
    tm_layout(legend.position = tm_pos_in("center", "bottom"),
              legend.orientation = "landscape",
              legend.bg.color = "white")
  
  plot_1 <- ggplot(data = clim_pts, aes_string(x = clim_var1, y = clim_var2)) +
    geom_point(shape = 16, color = "#3a5a40") +
    labs(title = paste("Species climate niche of", str_to_sentence(species_name))) +
    theme_bw()
  
  plot_2 <- ggplot(data = clim_random_pts, aes_string(x = clim_var1, y = clim_var2)) +
    geom_point(shape = 16) +
    labs(y = element_blank(), 
         title = paste("Background climate of", str_to_sentence(species_name))) +
    theme_bw()
  
  assign(paste0(species_name_lower, "_map_1"), map_1, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_map_2"), map_2, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_plot_1"), plot_1, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_plot_2"), plot_2, envir = .GlobalEnv)
  
}
```




